\thetitle

### 2009-02-24 paper/paper.tex
Yet more final tweaks.
Ignore-this: 200941a53a064f208d2b0bbcc35b4cd0

darcs-hash:20090224143438-bab43-74608dd14c2310bd1b6a55f84e6d46c6113f2607.gz

For reasons that will become clear in later sections, we compose the monad we will instantiate the above definitions with using a monad transformer \cite{monadtrans} |MMT| (for ``monoid monad transformer''), which piggybacks a monoid onto an existing monad by pairing.

@@

We now use |MMT| to piggyback the additive monoid on |nat| onto the identity monad, and lift elementwise comparison into the resulting monad, which we call |SP| (for ``simply-profiled'').
\begin{code}
  Definition SP: Monad := MMT NatAddMonoid IdMonad.

@@

To denote the expected value of a \emph{measure} of the output of a program, we define

@@

The last thing needed before the main theorem can be stated, is the notion of big-O complexity. We use a standard textbook definition, except that we incorporate a measure function parameter |m|:

@@

  Notation {-"``"-}over m,{-"\ "-} f =O( g ){-"\text{''''}"-} := bigO m f g.
\end{code}
We now state the main theorem.
\begin{code}
  Theorem qs_avg: over length, expec cost fcmp qs_NDP =O(fun n => INR (n * log2ceil n)).
\end{code}
Thanks to the property discussed at the start of the previous section, % perhaps too far back
|qs_avg| follows as a corollary from the stronger statement
\begin{code}
  qs_expec_cost: forall l, expec cost (qs_NDP l) <= 2 * length l * (1 + log2ceil (length l)),
\end{code}
the proof of which is described in the next two sections.

@@

\begin{code}
monoid_expec (count (i, j)) (qs_U (filter ((= Lt) fcmp lookup_cmp pivot) rest)) +
monoid_expec (count (i, j)) (qs_U (filter ((= Gt) fcmp lookup_cmp pivot) rest))
  <= ecc i j.
\end{code}
All indices in the first filtered list denote elements 
%%% that are 
less than the element denoted by the pivot. Since the former precede the latter in |sort l|, it must be the case that these indices are all less than |pivot|. And since |pivot < i|, it follows that the first |qs_U| term will produce no |(i, j)| comparisons (using property 
   (\ref{monoid_expec_zero_lem})  
shown at the end of section~\ref{reduction}). Hence, the first |monoid_expec| term vanishes%%% 
%%% as well
, leaving
\begin{code}
monoid_expec (count (i, j))
  (qs_U (filter ((= Gt) fcmp lookup_cmp pivot) rest)) <= ecc i j.
\end{code}
We now compare |nth (sort l) i| with |nth (sort l) pivot|.
\begin{itemize}
\item If the two are equal, then |i| will not occur in the |filter| term, and so (again) no |(i, j)| comparisons are performed.
\item If |nth (sort l) i < nth (sort l) pivot|, then we must have |i < pivot|, contradicting the assumption that |pivot < i|.
\item If |nth (sort l) i > nth (sort l) pivot|, then we apply the induction hypothesis. For this, it must be shown that filtering the list of indices preserves contiguity, which follows from the fact that the indices share the order of the elements they denote in |sort l|.
\end{itemize}
This concludes the case where |pivot < i|. The case where |j < pivot| is symmetric. The other three cases use similar arguments.
%%%
The proof is now complete.

>>>>

For reasons that will become clear in later sections, we construct the monad with which we will instantiate the above definitions using a monad transformer \cite{monadtrans} |MMT| (for ``monoid monad transformer''), which piggybacks a monoid onto an existing monad by pairing.

@@

We now use |MMT| to piggyback the additive monoid structure on |nat| onto the identity monad, and lift elementwise comparison into the resulting monad, which we call |SP| (for ``simply-profiled'').
\begin{code}
  Definition SP: Monad := MMT NatAddMonoid IdMonad.

@@

To denote the expected value of a discrete measure |f| of the output of a program, we define

@@

The last thing needed before the main theorem can be stated, is the notion of big-O complexity. We use the standard textbook definition, except that we make explicit how we measure inputs to |f|, namely with respect to a measure function |m|:

@@

  Notation {-"``"-}wrt m,{-"\ "-} f =O( g ){-"\text{''''}"-} := bigO m f g.
\end{code}
We now state the main theorem.
\begin{code}
  Theorem qs_avg: wrt length, expec cost fcmp qs_NDP =O(fun n => INR (n * log2ceil n)).
\end{code}
Thanks to the property discussed at the start of the previous section, % perhaps too far back
|qs_avg| follows as a corollary from the stronger statement
\begin{code}
  qs_expec_cost: forall l, expec cost (qs_NDP l) <= 2 * length l * (1 + log2ceil (length l)),
\end{code}
the proof of which is described in the next two sections.

@@

\begin{code}
monoid_expec (count (i, j)) (qs_U (filter ((= Lt) fcmp lookup_cmp pivot) rest)) +
monoid_expec (count (i, j)) (qs_U (filter ((= Gt) fcmp lookup_cmp pivot) rest))
  <= ecc i j.
\end{code}
All indices in the first filtered list denote elements 
%%% that are 
less than the element denoted by the pivot. Since the former precede the latter in |sort l|, it must be the case that these indices are all less than |pivot|. And since |pivot < i|, it follows that the first |qs_U| term will produce no |(i, j)| comparisons (using property 
   (\ref{monoid_expec_zero_lem})  
at the end of section~\ref{reduction} on page \pageref{monoid_expec_zero_lem}). Hence, the first |monoid_expec| term vanishes%%%
%%% as well
, leaving
\begin{code}
monoid_expec (count (i, j))
  (qs_U (filter ((= Gt) fcmp lookup_cmp pivot) rest)) <= ecc i j.
\end{code}
We now compare |nth (sort l) i| with |nth (sort l) pivot|.
\begin{itemize}
\item If the two are equal, then |i| will not occur in the |filter| term, and so (again) no |(i, j)| comparisons are performed.
\item If |nth (sort l) i < nth (sort l) pivot|, then we must have |i < pivot|, contradicting the assumption that |pivot < i|.
\item If |nth (sort l) i > nth (sort l) pivot|, then we apply the induction hypothesis. For this, it must be shown that filtering the list of indices preserves contiguity, which follows from the fact that the indices share the order of the elements they denote in |sort l|.
\end{itemize}
This concludes the case where |pivot < i|. The case where |j < pivot| is symmetric. The other three cases use similar arguments.
%%%
The proof is now complete.
